% \begin{document}
\chapter{Quantum Computing}
Whereas classical computers are based on bits that can either be switched "on" or "off", quantum computers work by carefully modifying the state of a quantum system. This fundamental difference in models of computation requires a complete remodelling of the mathematical representation that describes how a computer behaves.

\section{Qbits}
Qbits, short for quantum bits, are the fundamental element of quantum information. They are implemented as a binary property of a quantum system. This means that when said property is measured, the output is either "on" or "off", just like a classical bit. The difference is that quantum superposition allows the system to be in multiple states at the same time, and when the property is observed it stochasticaly collapses into a single base state. To model this crazy behaviour, the state of a qbit $ |\psi> $ can be represented as a linear combination:
\[
|\psi> = c_0 |0> + c_1 |1>
\]
where $ |0> $ and $ |1> $ indicate the "off" and "on" states and $ c_0, c_1 \in \mathbb{C} $ such that $ |c_0|^2 $ and $ |c_1|^2 $ are the probabilities of the qbit being in the relative state when measured. The reason for using the squared modulus of complex numbers as opposed to directly using probabilities as coefficients is that this allows for addition to lower the resulting probabilities. This property is needed to correctly model quantum interference, .....

So we can represent a qbit as a normalised vector in $ \mathbb{C}^2 $ with $ \begin{pmatrix}
1\\
0\\
\end{pmatrix} = |0> $ and $ \begin{pmatrix}
0\\
1\\
\end{pmatrix} = |1> $ as base:
\[
|\psi> = \begin{pmatrix}
c_0\\
c_1\\
\end{pmatrix}
\]
\section{Multi-Qbit Systems}
When adding a qbit to a quantum system, the number of possible base states doubles, meaning that an $ n $-qbit system's state is represented as a vector in $ \mathbb{C}^{2^n} $. Given two quantum systems, their composition is represented by the tensor product of their vector spaces. (give example and code)

The state of a multi-qbit system can't be described as the product of single-qbit states, as shown by the following example:

(bell pair example)

This mathematical fact describes a unique quantum behaviour known as entanglement, which is when two .....

\section{Quantum Gates}
Because we're representing a quantum state as a vector, any operator that acts on qbits will be an endomorfism, and thus a linear application with an associated square matrix $ U \in M(\mathbb{C})n\times n $.

As described in Shrodinger's equation (cite maybe?), a closed quantum system can only undergo unitary evolution, thus the only way we can we can manipulate our quantum state without it collapsing is through unitary matrices. A matrix is unitary if (define unitary matrix and how they're isometric)...

So all possible operators that act on a closed quantum system, known as quantum gates, are isomorfic to the group of unitary matrices of size $ n $. As a result, quantum gates are limited to only reversible operations that cannot clone a qbit's state to another without collapsing the original.

A set of gates is universal if, through composition, they can simulate any other quantum gate. The set of gates that will be used in this thesis is the following: (TODO: fix a gate set)

\section{Quantum Circuits}

A quantum program can be represented as a single unitary matrix acting on all qbits, depicted as follows:

(image)

where the wires represent the life span of the qbits. The single complex gate can be broken down into a composition of multiple simpler gates that are applied to a subsect of qbits, so we can deptict how the program builds the corresponding circuit sequencially:

(image)

We can define some structural proprerties of such a circuit:
\begin{itemize}
\item Gate count: 
\item Width: 
\item Depth:
\end{itemize}

\section{Circuit Optimization}
It's possible that the construction of the circuit determined by the quantum program isn't the most efficient decomposition, given a fixed gate set, of the ?? program matrix ?? with respect to some of the circuit's properties. This means that the circuit can be optimized. There are different ways a quantum circuit can be transformed into an equivalent, more optimized, one.

(describe different optimization techniques, ending with triviality conditions (the one ill use))

\chapter{Hoare Logic and Automated Verification}
Formally verifying a computer program means proving its correctness, with respect to a formal specification of the programming language used, by using formal or mathematical methods. This is the strongest possible way to determine if a certain block of code works as expected or not, and because of this formal verification is used mostly in security-related applications. For the purposes of this thesis, we are interested in such techniques as they don't require running or simulating quantum code in order to reason about its properties.

\section{Hoare Logic}
Hoare Logic is an axiomatic system for formally specifying what a program does. In 1969 C.A.R. Hoare introduced the Hoare Triple, written as follows:
\[
\{P\}C \{Q\}
\]
where:
\begin{itemize}
  \item $ C $ is a program (or any block or single command)
  \item $ P $ and $ Q $ are logical assertions, called conditions, describing the program's state immediatey before/after the execution of $ C $. P is known as the precondition and $ Q $ as the postcondition.
\end{itemize}

The conditions can be written using any well-defined formal logic that allows for the required descriptiveness, wich in our case is propositional logic (TODO: or first order?).

We will only be interested in the \textit{partial correctness} of our program: if the code in $ C $ loops indefinetly, then the triple is considered vacously true. \textit{Total correctness} is when, given $ P $, $ C $ is guaranteed to terminate and be in state $ Q $.

Hoare Logic constructs proofs by using axioms and inference rules that match with the syntactic elements of a programing language. Some of the core rules are as follows:
\begin{itemize}
\item \textbf{Rule of Composition}: allows sequential operations to be split into separate triples
\item \textbf{Assignment Axiom}: ensures that the postcondition $ Q  $ holds after a variable assignment
\item \textbf{If-Else Rule}:
\item \textbf{While Rule}:
\end{itemize}

While Hoare Logic provides a framework for verifying proofs, given a complete set of rules and axioms, it isn't well suited for proof automation. The following section will discuss how we can do just that. 

\section{Weakest Precondition Calculus}
In order to build a verification engine based of Hoare's framework, Edgar Djikstra transformed the logic into a backward-directed deterministic function. Given a piece of code $ C $ and a postcondition $ Q  $, 

\section{SMT Solvers}


% \end{document}
